Manuscript_draft

### 2021-03-24 manuscript_draft.Rmd
manny work

---
title: "Manuscript_draft"
author: "Michael Jull and Hardy Griesbauer"
date: "10/12/2020"
output:
  word_document:
    reference_docx: wordTemplate2.docx
  html_document:
    df_print: paged
csl: cjfr.csl
bibliography: library.bib
editor_options: 
  markdown: 
    wrap: 72
---

@@

# Load libraries
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom)
library(MuMIn)
library(ggeffects)
library(car)
library(flextable)
library(captioner)
library(lsmeans)
library(cowplot)
library(officer)
library(sjPlot)
library(emmeans)

@@

To test the hypothesis that tree-level radial growth response over the
1994-2019 period varied with interactions between (i) tree diameter at
the time of final treatment (1994); (ii) treatment; and (iii) species,
we fit linear mixed effects models with log-transformed 1994-2019
diameter increment as the response variable, and treatment, species and
log-transformed 1994 diameter as interacting fixed effects. To account
for plot-level productivity differences, we included plot as a random
intercept effect. Models were fit using the 'lme4' package for R [\@].
We also fit a similar model with the 1994-2019 diameter increment
expressed as a percentage of 1994 diameter. Plots of residuals and fit
lines with orginal data were examined to evaluate model goodness of fit.
Variance explained for each model are reported using the approach
described in Nakagawa et al. [-@Nakagawa2013], as implemented in the
MuMIn package for R [@barton2009mumin]. Pairwise contrasts were
performed using the 'emmeans' package for R [\@] and Tukey HSD method of
*p* value adjustment was used to control for multiplicity. We report all
significant results at p\<0.05.

For each plot and measurement period, we computed the growth dominance statistic (GDS) to evaluate the relationship between individual tree size and its growth as a proportion of the stand's overall growth [@Forrester; @West; @Binkey]

@@

The growth dominance coefficient was negative for all plots and time periods, indicating that smaller trees contributed
proportionally more to stand basal area growth than larger trees
(`r figNums("gds.plot",display="cite")`). Over the last two measurement
periods, the growth dominance coefficient increased in all plots except
one (not shown), indicating that individual tree growth contributions to
stand growth became more proportional to tree size over the last 11
growing seasons.

@@

## Uneven-aged stand management

This study describes tree- and stand-level responses to partial harvest
treatments in an uneven-aged stand.\
- THere are numerous studies documenting responses of even-aged stands
to partial harvest treatments such as thinning. In even-aged stands, a
common objective of thinning treatments is to re-allocate growing space
and resources to the most valulable trees in the stand that have
capacity to increase production. The objectives of partial harvesting in
uneven-aged stands is more often to provide a sustained production of
wood over multiple generations of trees. In this context, partial
harvesting in uneven-aged stands seeks to provide adequate growing space
and resources to trees across a range of sizes and densities. By
quantifying growth responses at the tree- and stand-level, this study
addresses questions regarding the manipulation of structure in an
uneven-aged stand to achieve ...

@@

At the tree-level, our results suggest that while both hybrid white
spruce and subalpine fir respond to thinning with faster radial growth,
heavy thinning can elicit a stronger radial growth response in spruce,
compared to fir. Spruce trees in the low-removal harvest treatment did
not respond with a significant growth release. Thus, our results suggest
that a relatively high removal of the stand will result in increased
spruce growth, and this release will exceed that of co-occurring fir
trees. Other studies have shown that residual spruce growth responds
favourably to heavy thinning. A 10-year analysis of white spruce growth
underneath an aspen canopy found that a complete or high overstory
removal resulted in immediate and significant spruce radial and height
growth responses, compared to control and partial cutting treatments
(Smith et al. 2016). In a 35-year old white spruce stand, diameter
increment was maximized with the heaviest thinning intensity, however,
the treatment thinning from below (Stiell 1970). Frank (1973) also
showed that basal area increment in a 70-year old white spruce stand increased with the
amount of thinning. Competition for light is a key factor that limits
white spruce growth, and opening up the stand to allow for increased
light penetration may be a primary mechanism that enhances white spruce
growth responses [@Lieffers1999]. High basal area removal may also
increase air and soil temperatures, which can preferentially favour
spruce growth responses compared to fir [@Lajzerowicz2004]. The
relatively low spruce growth response in the low-removal treatment may
be explained in that the gaps created may have been too small to elicit
as strong a growth response as observed in the high-removal thinning
treatment. Light levels may differ little between forest canopies that
are closed or contain small gaps [@Bartemucci2002], and strong vertical
light gradients may result in relatively little light regime changes for
understorey trees.

In comparison, both partial harvest treatments stimulated a growth
response in fir, compared to the control. Firs are considered to be better adapted than spruce to
low light conditions in the understory, allowing them to persist longer
in undisturbed forest (Antos 2000) and potentially able to respond
faster to even modest changes in understory light (Dumais et al 2014,
Messier et al. 1999). Although fir growth may respond at lower levels of
overstorey removal, the growth response potential of fir may be lower
than spruce. Comparative studies of spruce and fir growth rates present
differing evidence regarding genus-level differences in release potential to
thinning. In mature Engelmann spruce (*Picea engelmannii*) - subalpine
fir forests in western North America, understory spruce growth rates
exceeded those of fir, allowing spruce to achieve canopy position
faster, including after canopy gaps were created through disturbance
(Antos et al. 2000, Andrus et al. 2018). In a mature forest in central BC,
spruce growth exceeded that of subalpine fir after age 80 (Eis and
Craigdallie 1983). Because spruce tends to live longer than subalpine
fir, higher growth rates at later ages may allow this species to become
larger than fir. In contrast, thinning studies in spruce-fir forests in
eastern North America report that balsam fir (*Abies balsamifera*)
responds to release with faster height and diameter growth than
co-occurring red spruce (*Picea rubens*) (Dumais and Prevost, 2007;
Dumais et al. 2014, Messier et al. 1999). Other studies have found no
difference in growth response between spruce and fir species after
treatment (MCaughey and Smith1982). Our study also suggests that at
lower levels of basal area removal, spruce and fir growth responses are
similar.

Post-treatment growth rates increased with tree size at the time of
treatment, for both species. This has been found in other studies as
well (references) and may be related to size-asymmetric competition
whereby larger trees preempt resources, especially light, from smaller
individuals [@Bose2017]. Thinning treatments may exacerbate
size-assymetric competition [@Cordonnier2015], and growth increases from
treatment will be predominantly concentrated in the larger tree size
classes [@Bose2018; @Schwinning1998]. Smaller trees will likely continue
to experience competition and will not benefit as much from the thinning
treatment. Thinning treatments may not promote growth responses in the
largest dominant trees in a stand either, as these trees may not
experience significant changes in resource availability or may have
achieved maximal growth rates [@Ryan1997, Bose2017].

Although growth responses were concentrated in the larger trees, our analysis of growth dominance indicated that smaller trees contributed proportionally more to stand growth than larger trees.  Binkley et al. (2006) propose that GDC in developing forests increases from 0 to positive during early development, reflecting the growth dominance of larger robust trees as canopy closure is achieved.  As the forests age, the GDC may switch to negative, reflecting several co-occuring processes.  Older trees may start to senesce, with growth reductions and increased mortality.  Younger trees may increase their resource use efficiency, and/or increase their allocation to leaf biomass (Baret et al. 2017) partially in response to increasing resource availiabilty as older trees senesce.  Data from old Englemann spruce – subalpine fir forests in Colorado (Binkley et al. 2006) as well as old-growth boreal mixedwood stands (Baret et al. 2017) support this proposed model of growth dominance development.  While we found a similar negative growth dominance in the stands of this study, a comparison between the two recent measurement periods showed that growth dominance actually increased over time (i.e., tree growth became increasingly proportional to tree size).  This finding is unique from other studies, and suggests that either larger trees are increasing their growth rates relative to smaller trees, or smaller trees are beginning to experience increased competition and may be experiencing lower growth rates.  Further data from these stands will help evaluate the long-term trend in GDC.

@@

While our study found that high-removal partial harvest resulted in
faster basal area growth, we did not find evidence that thinning changed
volume increment in trees greater than 17.5cm DBH. A possible
explanation is that tree growth responses to thinning might be primarily
concentrated in radial growth, as opposed to height growth; this would
be important because the volume formula used in this study is based on
both tree diameter and height [@Nigh2016]. We note that it is often
difficult to precisely capture height growth changes in mature trees
using conventional field-based methods [@Vepakommaetal2011], which makes
capturing height changes problematic in these sorts of studies. We also
utilized a 17.5cm DBH limit for volume calculations; as such, our volume
measures only reflect the contribution of the larger trees in the stand,
and do not include volume increment in smaller trees.

The partial harvest treatment appeared to change the timing of peak
stand growth. In the control unit, Volume and basal area increment
peaked in the 1994-1997 measurement period, and declined thereafter,
with the strongest decline in annual growth over the last measurement
period of 2009-2019. Age-related declines in stand productivity may be
associated with decreasing water-use efficiency and reduced
photosynthesis, with the age of maximal growth rate varying with factors
such as species and site conditions [@Ryan1997]. In both thinned
treatments, maximal growth rates peaked in the 1997-2009 period. Thus,
it appears the thinning treatment may have delayed the period of maximal
growth rates by approximately 10 years. The timing of thinning
treatments can be important. Stands that are past their period of
age-related maximal growth rate may not have as much capacity to respond
to treatments as a younger stand, although this is liekly a more
important factor for shade-intolerant species rather than the spruce and
fir trees in this study [@Bose2017]. At the time of this study, the
high-removal thinning treatment had maintianed an annual basal area and
volume increment that was similar to previous measurement periods; if
this high level of increment continues, differences in basal area and
volume will continue to diminish compared to the control, and this stand
may exceed the low-removal thinning treatment.

>>>>

---
title: "Manuscript_draft"
author: "Michael Jull and Hardy Griesbauer"
date: "10/12/2020"
output:
  word_document:
    reference_docx: wordTemplate2.docx
  html_document:
    df_print: paged
csl: cjfr.csl
bibliography: C:\\Users\\hgriesba\\Google Drive\\Papers\\library.bib
editor_options: 
  markdown: 
    wrap: 72
---

@@

# Load libraries
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom)
library(MuMIn)
library(ggeffects)
library(car)
library(flextable)
library(captioner)
library(lsmeans)
library(cowplot)
library(officer)
library(emmeans)

@@

To test the hypothesis that tree-level radial growth response over the
1994-2019 period varied with interactions between (i) tree diameter at
the time of final treatment (1994); (ii) treatment; and (iii) species,
we fit linear mixed effects models with log-transformed 1994-2019
diameter increment as the response variable, and treatment, species and
log-transformed 1994 diameter as interacting fixed effects. To account
for plot-level productivity differences, we included plot as a random
intercept effect. Models were fit using the 'lme4' package [@Bates2015] for the R statistical environment [@R-base].
We also fit a similar model with the 1994-2019 diameter increment
expressed as a percentage of 1994 diameter. Plots of residuals and fit
lines with orginal data were examined to evaluate model goodness of fit.
Variance explained for each model are reported using the approach
described in Nakagawa et al. [-@Nakagawa2013], as implemented in the
MuMIn package for R [@barton2009mumin]. Pairwise contrasts were
performed using the 'emmeans' package for R [@Lenth2019] and Tukey Honestly Significant Difference test of
*p* value adjustment was used to control for multiplicity. We report all
significant results at p<0.05.

Growth dominance in a stand considers the contribution of individual trees to total stand growth as a function of their size [@Binkley2006].  The growth dominance statistic [GDS; @Binkley2006; @West2014] represents growth dominance as a single metric, whereby a GDS value of 0 indicates a stand where the growth of individual trees, expressed as a proportion of stand total growth, is equal to their size, expressed as a proportion of stand total size (e.g., basal area).  A negative (positive) GDS indicates a stand where smaller (larger) trees contribute more to stand growth relative to their proportional size.  For formulae used to compute GDS, we refer the reader to West [-@West2014].

@@

The growth dominance coefficient was negative for all plots and time periods, indicating that smaller trees contributed
proportionally more to stand basal area growth than larger trees
(`r figNums("gds.plot",display="cite")`). Over the last two measurement
periods of 1997-2009 and 2009-2019, the growth dominance coefficient increased in all plots except
one (not shown), indicating that individual tree growth contributions to
stand growth became more proportional to tree size over the last 11
growing seasons.

@@

## Uneven-aged stand management
The successful application of partial harvesting in uneven-aged, multi-species stands can be complicated [@smith1997practice], especially where the objectives include the sustained production of wood over multiple generations of trees.  In this context, partial harvesting in uneven-aged stands seeks to re-allocate growing space and resources to trees across a range of sizes and densities.  Selection thinning is a typical thinning treatment in uneven-aged stands, whereby individual or groups of dominant trees are removed to recover their economic value as well as create growing space for understory trees.  While this treatment

@@

At the tree-level, our results suggest that while both hybrid white
spruce and subalpine fir respond to thinning with faster radial growth,
heavy thinning can elicit a stronger radial growth response in spruce,
compared to fir. Spruce trees in the low-removal harvest treatment did
not respond with a significant growth release. Thus, our results suggest
that a relatively high removal of the stand will result in increased
spruce growth, and this release will exceed that of co-occurring fir
trees. Other studies have shown that residual spruce growth responds
favourably to heavy thinning. A 10-year analysis of white spruce growth
underneath an aspen canopy found that a complete or high overstory
removal resulted in immediate and significant spruce radial and height
growth responses, compared to control and partial cutting treatments
[@smith2016sprucing]. In a 35-year old white spruce stand, diameter
increment was maximized with the heaviest thinning intensity, however,
the treatment thinning from below [@Stiell1980]. Frank [-@frank1973course] also
showed that basal area increment in a 70-year old white spruce stand increased with the
amount of thinning. Competition for light is a key factor that limits
white spruce growth, and opening up the stand to allow for increased
light penetration may be a primary mechanism that enhances white spruce
growth responses [@Lieffers1999]. High basal area removal may also
increase air and soil temperatures, which can preferentially favour
spruce growth responses compared to fir [@Lajzerowicz2004]. The
relatively low spruce growth response in the low-removal treatment may
be explained in that the gaps created may have been too small to elicit
as strong a growth response as observed in the high-removal thinning
treatment. Light levels may differ little between forest canopies that
are closed or contain small gaps [@Bartemucci2002], and strong vertical
light gradients may result in relatively little light regime changes for
understorey trees.

In comparison, both partial harvest treatments stimulated a growth
response in fir, compared to the control. Firs are considered to be better adapted than spruce to
low light conditions in the understory, allowing them to persist longer
in undisturbed forest [@Antos2000] and potentially able to respond
faster to even modest changes in understory light [@Dumais2014; @Messier1999]. Although fir growth may respond at lower levels of
overstorey removal, the growth response potential of fir may be lower
than spruce. Comparative studies of spruce and fir growth rates present
differing evidence regarding genus-level differences in release potential to
thinning. In mature Engelmann spruce (*Picea engelmannii*) - subalpine
fir forests in western North America, understory spruce growth rates
exceeded those of fir, allowing spruce to achieve canopy position
faster, including after canopy gaps were created through disturbance
[@Antos2000; @Andrus2018]. In a mature forest in central BC,
spruce growth exceeded that of subalpine fir after age 80 [@eis1983regeneration]. Because spruce tends to live longer than subalpine
fir, higher growth rates at later ages may allow this species to become
larger than fir. In contrast, thinning studies in spruce-fir forests in
eastern North America report that balsam fir (*Abies balsamifera*)
responds to release with faster height and diameter growth than
co-occurring red spruce (*Picea rubens*) [@Dumais2008; @Dumais2008a; @Dumais2014; @Messier1999]. Other studies have found no
difference in growth response between spruce and fir species after
treatment [@mccaughey1982understory]. Our study also suggests that at
lower levels of basal area removal, spruce and fir growth responses are
similar.

Post-treatment growth rates increased with tree size at the time of
treatment, for both species. This has been found in other studies as
well (references) and may be related to size-asymmetric competition
whereby larger trees preempt resources, especially light, from smaller
individuals [@Bose]. Thinning treatments may exacerbate
size-assymetric competition [@cordonnier2015gini], and growth increases from
treatment will be predominantly concentrated in the larger tree size
classes [@Bose; @schwinning1998mechanisms]. Smaller trees will likely continue
to experience competition and will not benefit as much from the thinning
treatment. Thinning treatments may not promote growth responses in the
largest dominant trees in a stand either, as these trees may not
experience significant changes in resource availability or may have
achieved maximal growth rates [@ryan1997age, @Bose].

Although growth responses were concentrated in the larger trees, our analysis of growth dominance indicated that smaller trees contributed proportionally more to stand growth than larger trees.  Binkley et al. [-@Binkley2006] propose that GDC in developing forests increases from 0 to positive during early development, reflecting the growth dominance of larger robust trees as canopy closure is achieved.  As the forests age, the GDC may switch to negative, reflecting several co-occuring processes.  Older trees may start to senesce, with growth reductions and increased mortality.  Younger trees may increase their resource use efficiency, and/or increase their allocation to leaf biomass [@baret2017long] partially in response to increasing resource availiabilty as older trees senesce.  Data from old Englemann spruce – subalpine fir forests in Colorado [@Binkley2006] as well as old-growth boreal mixedwood stands [@baret2017long] support this proposed model of growth dominance development.  While we found a similar negative growth dominance in the stands of this study, a comparison between the two recent measurement periods showed that growth dominance actually increased over time (i.e., tree growth became increasingly proportional to tree size).  This finding is unique from other studies, and suggests that either larger trees are increasing their growth rates relative to smaller trees, or smaller trees are beginning to experience increased competition and may be experiencing lower growth rates.  Further data from these stands will help evaluate the long-term trend in GDC.

@@

While our study found that high-removal partial harvest resulted in
faster basal area growth, we did not find evidence that thinning changed
volume increment in trees greater than 17.5cm DBH. A possible
explanation is that tree growth responses to thinning might be primarily
concentrated in radial growth, as opposed to height growth; this would
be important because the volume formula used in this study is based on
both tree diameter and height [@nigh2016]. We note that it is often
difficult to precisely capture height growth changes in mature trees
using conventional field-based methods [@vepakomma2011response], which makes
capturing height changes problematic in these sorts of studies. We also
utilized a 17.5cm DBH limit for volume calculations; as such, our volume
measures only reflect the contribution of the larger trees in the stand,
and do not include volume increment in smaller trees.

The partial harvest treatment appeared to change the timing of peak
stand growth. In the control unit, Volume and basal area increment
peaked in the 1994-1997 measurement period, and declined thereafter,
with the strongest decline in annual growth over the last measurement
period of 2009-2019. Age-related declines in stand productivity may be
associated with decreasing water-use efficiency and reduced
photosynthesis, with the age of maximal growth rate varying with factors
such as species and site conditions [@ryan1997age]. In both thinned
treatments, maximal growth rates peaked in the 1997-2009 period. Thus,
it appears the thinning treatment may have delayed the period of maximal
growth rates by approximately 10 years. The timing of thinning
treatments can be important. Stands that are past their period of
age-related maximal growth rate may not have as much capacity to respond
to treatments as a younger stand, although this is liekly a more
important factor for shade-intolerant species rather than the spruce and
fir trees in this study [@Bose]. At the time of this study, the
high-removal thinning treatment had maintianed an annual basal area and
volume increment that was similar to previous measurement periods; if
this high level of increment continues, differences in basal area and
volume will continue to diminish compared to the control, and this stand
may exceed the low-removal thinning treatment.
